$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-15 paper/paper.tex
Add work by Ian Mackie
pass on the intro

We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, build, and predict such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be transmitted.
The \picalc{} is a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and (more recently) Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types.
These make the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.

In the early '90s, \emph{linear types} for (functional) programming languages \cite{Wadler90,Bernardy2018} were introduced, following J.-Y. Girard's development of linear logic \cite{Girard87}.
Linear type systems guarantee that resources are used \emph{exactly once}.
Enforcing the ``no duplicating'' and ``no discarding'' of resources via linearity allows for resource-aware programming and more efficient implementations \cite{Wadler90}.
Later on, linearity inspired unique types (as in Clean \cite{BarendsenS96}) and ownership types (as in Rust \cite{MatsakisK14}).

Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98} (a formalism used in communication-centric programming), where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself is used multiple times, as per its session type.
Linearity in session types is the key ingredient that guarantees type safety.

Recently, an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} into the spotlight again \cite{KPT96}.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but is useful in practice too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}, and it allows to use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be built.

\emph{We present the first full mechanisation in Agda of a \picalc{} with linear, graded and shared types, all under a common unified framework.}

We do so by defining a specification for an algebra on \emph{multiplicities} (how many times a channel can be used) and use it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be mixed in the same type system.
Ultimately, we can exploit this formalisation as a unified framework for type-safe distributed modelling and programming using a variety of type systems, and build on top of it more advanced types, theories and languages.

\begin{example}[Example process]
\label{example-process}

We present an example that highlights the ways in which graded types can be used to prevent mistakes in the definition of \picalc{} processes.
The syntax uses de Bruijn indices for variables and is introduced in detail in \autoref{syntax}.
In short: $\new{} P$ creates a new channel and makes it available at index $0$ in the continuation $P$; $\comp{P}{Q}$ composes $P$ and $Q$ in parallel; $\recv{x}{}P$ receives along channel $x$ and makes the data available at index $0$ in the continuation $P$; and $\send{x}{y}{P}$ sends $y$ over channel $x$ and continues with $P$.
Both variable references and processes are indexed by the number of free variables they can refer to (for any natural number $n$, $\Var_n$ and $\Process_n$, respectively).
For the sake of clarity, we define the example subprocesses separately and parametrise them by the channels on which they operate.

\begin{figure}[h]
  \tiny
  \centering
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-y) at (2,2)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=green] (recv-z) at (4,2)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=brown,align=center] (sync) at (2,0)   {$\func{sync}$\\$ \, x \, y \, z$};

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{Four communicating example processes composed in parallel.}
  \label{fig:example-process}
\end{figure}

\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
  & \\
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{\suc c}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
  & \func{sync} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
  & \func{sync} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0}{} \\
  & && \recv{(\suc \; in1)}{} \\
  & && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
  & \func{example} && : \; \Process_{\constr{0}} & \\
  & \func{example} && = \new{} ( \func{send} \; \constr{0} \\
  & && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
  & && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
  & && \comp{}{ \func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}

The example contains four processes that are composed in parallel: two of them send data over a channel; one receives data twice; the fourth sticks together and synchronizes the previous three processes (\autoref{fig:example-process}).
The process doing the synchronisation first receives on both input channels and then sends the received data along the output channel.
Thanks to linear and graded types, the processes below can have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{sync}$ consumes one input multiplicity on the channels shared with $\func{send}$, and two output multiplicities on the channel shared with $\func{recv}$.
We use these types to give a suitable typing derivation in \autoref{example-derivation} (\autoref{leftover-typing}).

@@

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we follow Allais \cite{Allais2018a} and apply leftover typing to the \picalc{}.
We generalise the usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work we too are able to talk about fully consumed resources --- e.g., we can transmit $\lz$ multiplicities of a fully exhausted channel.

>>>>

We live in a concurrent and distributed world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, build, and predict such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be sent and received.
The \picalc{} is at the basis of the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and more recently Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties, such as deadlock freedom, livelock freedom or termination, and most notably, session types.
These make the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.

In the early '90s, \emph{linear types} for (functional) programming languages \cite{Wadler90,Bernardy2018} were introduced, following J.-Y. Girard's development of linear logic \cite{Girard87}.
Linear type systems guarantee that resources are used \emph{exactly once}.
Enforcing this ``no duplicating'' and ``no discarding'' of resources via linearity allows for resource-aware programming and more efficient implementations \cite{Wadler90}.
Later on, linearity inspired unique types, as in Clean \cite{BarendsenS96} and ownership types, as in Rust \cite{MatsakisK14}.

Linearity also influenced the $\pi$-calculus.
Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98}, a formalism used in communication-centric programming, where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself may be used multiple times, in particular by following its session type.
Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.

In recent years, an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} \cite{KPT96} into the spotlight again.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but is useful in practice too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}, thus allowing to use the target linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be defined and implemented.

In this paper we present \emph{the first full mechanisation in Agda of a \picalc{} with linear, graded and shared types, all under a common unified framework.}

While linear types are useful for resource-aware programming, graded and shared types are useful for a more flexible and 
mainstream programming.

We define a specification for an algebra on \emph{multiplicities} --- how many times a channel can be used --- and use it to apply \emph{leftover typing} for the first time to the \picalc{}, along the lines of Allais \cite{Allais2018a}.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be mixed in the same type system.
Ultimately, we can exploit this formalisation as a unified framework for type-safe distributed modelling and programming using a variety of type systems, and build on top of it more advanced types, theories and languages.

\begin{example}[The courier system]
\label{example-process}

We present an example that highlights the ways in which linear, graded and shared types can be used to prevent errors in the definition of \picalc{} processes.

A courier system consists of three roles: a sender, who wants to send a package; a receiver, who receives the package sent by the sender and a courier, who carries the package from the sender to the receiver.
Our courier system consists of four \picalc{} processes composed in parallel:
two sender processes, $S_1$ and $S_2$, sending data over channels $x$ and $y$, respectively;
one receiver process, $R$, which receives the data from each of the senders;
and finally a courier process $C$, which synchronises communication among the previous three process and
first receives data from $S_1$ and $S_2$ on its input channels, respectively $x$ and $y$ and then sends the received data to $R$ along its output channel $z$.
A graphical representation of the courier system is given in \autoref{fig:example-process}.

Let us now consider the $\pi$-calculus process behind the courier system.
The syntax of \picalc{} processes adopts de Bruijn indices \cite{deBruijn1972} for channel names and variables and is introduced in detail in \autoref{syntax}.
In short, for the purposes of this example:
process $\new{} P$ creates a new channel and makes it available at index $0$ in the continuation process $P$;
$\comp{P}{Q}$ composes processes $P$ and $Q$ in parallel; process $\recv{x}{}P$ receives data along channel $x$ and makes that data available at index $0$ in the continuation process $P$;
process $\send{x}{y}{P}$ sends variable $y$ over channel $x$ and continues as process $P$.
Both variable references and processes are indexed by the number of free variables they can refer to, respectively, $\Var_n$ and $\Process_n$, for any natural number $n$.

For the sake of clarity, we will define the four subprocesses in our system separately and parametrise them by the channels on which they operate.

\begin{figure}[t]
  \small
  \centering
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-y) at (2,2)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=green] (recv-z) at (4,2)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=blue,align=center] (sync) at (2,0)   {$\func{cour}$\\$ \, x \, y \, z$};

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{The courier system: two senders $S_1$ and $S_2$, one receiver $R$ and one courier $C$.}
  \label{fig:example-process}
\end{figure}

The receiver process $R$ is defined as follows:
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
\end{flalign*}

The sender processes sender processes $S_1$ and $S_2$ are defined as follows:
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{\suc c}{\constr{0}} \PO \\
\end{flalign*}

The courier process $C$ is defined as follows:
\begin{flalign*}
  & \func{cour} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
  & \func{cour} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0}{} \\
  & && \recv{(\suc \; in1)}{} \\
  & && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
  & \func{courier} && : \; \Process_{\constr{0}} & \\
  & \func{courier} && = \new{} ( \func{send} \; \constr{0} \\
  & && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
  & && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
  & && \comp{}{ \func{cour} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}
[ALIGN FIG AND PROCESSES IN PARTICULAR send c and send x,y in THE FIGURE]
[COMMENT ON THE PROCESSES AND WHAT THEY DO]
[EXPLAIN HERE WHAT IS TYPED WITH LIN, GRADED AND SHARED]
Thanks to linear and graded types, the above processes can have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{cour}$ consumes one input multiplicity on the shared channels with $\func{send}$, respectively $x$ and $y$ in \autoref{fig:example-process} and two output multiplicities on the channel shared with $\func{recv}$, channel $z$ in \autoref{fig:example-process}.
We use these types to give a suitable typing derivation in \autoref{example-derivation} (\autoref{leftover-typing}).
\end{example}

@@

To the best of our knowledge, leftover typing makes its appearance in 1994, when Ian Mackie first uses it to formulate intuitionistic linear logic \cite{Mackie}.
Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we follow Allais \cite{Allais2018a} and apply leftover typing to the \picalc{}.
We generalise the usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work we too are able to talk about fully consumed resources --- e.g., we can transmit $\lz$ multiplicities of a fully exhausted channel.
